---
title: "mappingAS: IUCN Criterion B screening (EOO, AOO and habitat conversion)"
author: "mappingAS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mappingAS: IUCN Criterion B screening (EOO, AOO and habitat conversion)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE        # examples read MapBiomas over the network; not run here
)
```

## What the package does

`mappingAS` is a [GeoCAT](https://geocat.iucnredlist.org/)-style tool for
**preliminary IUCN Red List Criterion B screening**. From occurrence points
(`.xlsx`, `.csv` or a spatial file) it:

- computes the **EOO** (Extent of Occurrence, minimum convex polygon) and the
  **AOO** (Area of Occupancy, 2 km grid) on a data-centred equal-area
  projection, and assigns a provisional Criterion B category;
- measures the **% converted (anthropic)** and **% natural** area within the EOO
  and the AOO, plus the **area of each land-cover class**, from MapBiomas (or, for
  ranges outside its coverage, the global Sentinel-2/Esri 10 m product);
- builds the land-cover **time series** (% by year);
- derives the Red List **supporting analyses** described below (countries of
  occurrence, Area of Habitat and the lower/upper AOO bounds, elevation
  preferences, severe-fragmentation screening);
- writes a narrative **assessment report**, and exports everything (tables,
  shapefile/GeoPackage). A **Shiny app** ties it all together.

> The categories are **provisional** (EOO/AOO size thresholds only) and are for
> screening, not a formal Red List assessment.

```{r load}
library(mappingAS)
```

## 1. Import the points

`read_occurrences()` accepts `.csv`/`.tsv`/`.txt`, `.xlsx`/`.xls` and spatial
files (`.shp`, `.gpkg`, `.geojson`, `.zip`) and tries to **detect** the species,
longitude and latitude columns automatically.

```{r import}
ex  <- system.file("extdata", "example_occurrences.csv", package = "mappingAS")
occ <- read_occurrences(ex)

# if detection fails, name the columns explicitly:
# occ <- read_occurrences("my_data.csv",
#                         species_col = "species",
#                         lon_col = "longitude", lat_col = "latitude")
```

The result is an `sf` object of points in WGS84 with a `species` column.

## 2. Assess the species

`assess_species()` runs the whole chain per species: EOO, AOO, conversion and the
provisional category.

```{r assess}
res <- assess_species(
  occ,
  year = 2024,          # MapBiomas year for the land-cover snapshot
  collection = 10,      # MapBiomas collection
  backend = "local"     # "local" (no GEE account) or "gee"
)

res            # compact print
res$summary    # one row per species
```

Key `res$summary` columns: `eoo_km2`, `aoo_km2`, `aoo_cells`,
`eoo_converted_pct`/`eoo_natural_pct`, `aoo_converted_pct`/`aoo_natural_pct`,
`provisional_cat` and `category_B`. The conversion index uses a **terrestrial**
denominator: `converted = anthropic / (anthropic + natural) * 100` (water and
"not observed" are excluded by default; change with
`water_in_denominator = TRUE`).

## 3. Map and conversion chart

```{r viz}
map_species(res)            # leaflet: points + EOO + AOO
plot_conversion(res)        # bars: natural / altered / water / other
```

## 4. Per-class composition

```{r classes}
ct <- class_table(res)
head(ct)
# species | range (EOO/AOO) | code | class_pt | class_en | group | area_km2 | pct
```

## 5. Land-cover time series

```{r timeseries}
# every 5 years, within the first species' EOO
ts <- timeseries_for_species(res, range = "eoo", by = "class",
                             years = c(2020, 2024))
plot_timeseries(ts)         # stacked-area chart, official MapBiomas colours

# or over any geometry:
# ts2 <- cover_timeseries(my_geometry, years = c(1990, 2000, 2010, 2020),
#                         by = "class", backend = "local")
```

Each year is read separately from MapBiomas, so a full annual series can be slow;
the minimum reliable step is 5 years.

## 6. Red List supporting analyses

These assimilate the [sRedList](https://sredlist.eu/) workflow (Cazalis *et al.*
2024) while keeping the focus on Criterion B, computed from the data the package
already produces. Most take the range polygon; take it from `calc_eoo()`.

```{r supporting}
sp1  <- occ[occ$species == occ$species[1], ]
hull <- calc_eoo(sp1)$hull          # EOO polygon
aoo_lo <- calc_aoo(sp1)$area_km2     # occurrence-based (lower) AOO

# Countries of occurrence + biogeographical realm (needs 'rnaturalearth'):
countries_of_occurrence(range = hull, points = sp1)

# Area of Habitat (suitable land cover within the EOO, optionally masked to an
# elevation band; needs land cover, and 'elevatr' or a DEM for elevation):
aoh <- calc_aoh(hull, year = 2024, elev_min = 0, elev_max = 1500, points = sp1)
aoh$aoh_km2            # Area of Habitat
aoh$aoo_upper_km2      # habitat-based (upper) AOO bound, at the 2 km scale
aoh$point_prevalence  # share of records inside the habitat (a validation metric)

# Suggested elevation band from the occurrences:
elevation_preferences(sp1)

# Lower/upper AOO bounds and the resulting B2 category range (Brooks et al. 2019):
b <- aoo_bounds(aoo_lower_km2 = aoo_lo, aoh_km2 = aoh$aoo_upper_km2,
                eoo_km2 = calc_eoo(sp1)$area_km2)
b$cat_range
iucn_category_B_range(eoo_km2 = calc_eoo(sp1)$area_km2,
                      aoo_lower_km2 = b$aoo_lower, aoo_upper_km2 = b$aoo_upper)

# Severe-fragmentation screening (Santini et al. 2019): cluster occurrences by a
# dispersal / isolation distance and see how much of the population is in small
# subpopulations. It never sets sub-criterion (a) automatically.
assess_fragmentation(sp1, isolation_km = 20)
```

## 7. Narrative report

`assessment_report()` writes an interpretive Criterion B summary as HTML, plain
text or a Word `.docx` file.

```{r report}
cat(assessment_report(res, output = "text"))

# Word document (needs the 'officer' package):
# assessment_report(res, output = "docx", file = "assessment.docx")
```

## 8. Export

```{r export}
export_ranges(res, dir = "out")                 # two shapefiles (EOO, AOO) + CSV
export_ranges(res, dir = "out", format = "gpkg") # one GeoPackage (eoo, aoo layers)
export_ranges(res, dir = "out", zip = TRUE)      # everything in one .zip
```

The attribute table carries `species`, `eoo_km2`/`aoo_km2`, `conv_pct`,
`nat_pct`, `cat_B1`/`cat_B2`, `prov_cat`, `mb_year` and `mb_coll`; the
`*_classes.csv` file gives each class's area and percentage.

## 9. Shiny app

```{r app}
run_app()
```

The app has tabs for the **Map**, **Add points**, **Conversion**, **Classes**,
**Protected areas**, **Time series**, **Fire**, **Countries**, **Habitat**,
**Fragmentation**, and, at the end of the workflow, **Assessment**, **Report**
and **Methods** - with CSV, shapefile/GeoPackage and image (PNG/HTML) downloads.

## 10. MapBiomas backends

- **`"local"` (default)**: reads only the *window* of the national GeoTIFF over
  the network with GDAL `/vsicurl/`. No Earth Engine account and no full-mosaic
  download. For offline/repeated use, point `src=` at a local GeoTIFF.
- **`"gee"`**: computes the per-class areas server-side on Google Earth Engine
  (recommended for very large ranges). Requires `rgee`:

```{r gee}
# install.packages("rgee"); rgee::ee_install(); rgee::ee_Initialize()
res <- assess_species(occ, backend = "gee")
```

## Caveats

- The **categories are provisional** (size thresholds only) - do not report them
  as final categories. The supporting analyses (fragmentation, decline, the
  AOO bounds) inform the judgement but the assessor decides.
- The **AOO** is sensitive to grid placement and sampling effort.
- **MapBiomas accuracy** varies by class, biome and year.
- For continental ranges, prefer `backend = "gee"`.

## Citation

Please also cite **MapBiomas** (the collection used), the **IUCN Red List
Guidelines** for Criterion B, **GeoCAT** (Bachman *et al.* 2011, ZooKeys), and,
for the supporting analyses, **sRedList** (Cazalis *et al.* 2024, *Biological
Conservation*).
